---
title: "p8105_hw3_sx2402"
author: "Eric Xu"
date: "2025-10-04"
output: github_document
---
```{r settings,include=FALSE}
library(p8105.datasets)
library(tidyverse)
data("instacart")
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

## Problem 1:
```{r skim, eval = FALSE}
instacart_skim = instacart |> 
  mutate(
    user_id = as.character(user_id),
    aisle_id = as.character(aisle_id),
    reordered = as.factor(reordered),
    order_id = as.character(order_id),
    product_id = as.character(product_id),
    department_id = as.character(department_id)
         )

names(instacart_skim)
skimr::skim(instacart_skim)
```
There are 1384617 observations from dataset `instacart` and 131,209 unique users. Each row represents a single product in a customer’s order. There are in total 39123
unique product names. There are total 15 variables, 5 describing identifiers of order, aisle, product, department, and user; and `add_to_cart_order` shows order in which each product was added to cart.


Some other important variables include:

* `redordered`: if this product has been ordered by this user in the past
* `order_number`: the order sequence number for this user (1=first, n=nth)
* `order_dow`: the day of the week on which the order was placed
* `order_hour_of_day`: the hour of the day on which the order was placed
* `days_since_prior_order`: days since the last order, capped at 30, NA if order_number is 1

### Questions:

__How many aisles are there, and which aisles are the most items ordered from?__
```{r Q1, collapse=TRUE}
instacart_skim = instacart |> 
  summarize(
    n_aisle = n_distinct(aisle),
    n_obs = n()
    )
instacart_skim

instacart_skim = instacart |>
  group_by(aisle) |> 
  count(aisle, name = "n_times") |> 
  arrange(desc(n_times))
instacart_skim
```

_There are 134 unique aisles and aisle `fresh vegetables` occurred 150609 times with the the most items ordered from_

__Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.__
```{r Q2,fig.width=8, fig.height=6, dpi=250}
aisle_counts = instacart |> 
  count(aisle, name = "n_times") |> 
  filter(n_times > 10000)

aisle_counts |> 
  mutate(aisle = fct_reorder(aisle, -n_times)) |> 
  ggplot(aes(x = aisle, y = n_times)) +
  geom_bar(stat = "identity")+
   viridis::scale_color_viridis()+
  theme(axis.text.x = element_text(angle=90, size=9, hjust=1, vjust=0.5))+
  labs(
    title = "Number of items ordered by aisle",
    x = "Aisle",
    y = "Number of items ordered"
  )
```

__Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.__

```{r Q3}
top_items = instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |> 
  count(aisle,product_name, name = "times") |>
  group_by(aisle) |> 
  arrange(desc(times)) |> 
  slice_max(times, n =3)
knitr::kable(top_items)
```


__Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).__
```{r Q4}
mean_hours = instacart |> 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |> 
  group_by(product_name, order_dow) |> 
  summarize(
    mean_hour = mean(order_hour_of_day)
    ) |> 
  mutate(order_dow = case_when(
    order_dow == 0 ~ "Sunday",
    order_dow == 1 ~ "Monday",
    order_dow == 2 ~ "Tuesday",
    order_dow == 3 ~ "Wednesday",
    order_dow == 4 ~ "Thursday",
    order_dow == 5 ~ "Friday",
    order_dow == 6 ~ "Saturday"
  )) |> 
  pivot_wider(names_from = order_dow, values_from = mean_hour) |> 
  rename("Product Name" = "product_name")


knitr::kable(mean_hours, digits = 2)
```

## Problem 2
